\makeatletter \@ifundefined{rootpath}{\input{../../setup/preamble.tex}}\makeatother
%\worksheetstart{Titlepage}{0}{December 31, 2012}{../../}
\begin{titlingpage}
\aautitlepage{%
  \englishprojectinfo{
    Investigation of trending concurrency models by comparison of performance and characteristics: Software Transactional Memory, Actor Model, and Threads \& Locks. %title
  }{%
    Fall Semester 2014 %project period
  }{%
    dpt907e14  % project group
  }{%
    %list of group members
    Tobias Ugleholdt Hansen\\
    Andreas Pørtner Karlsen\\ 
    Kasper Breinholt Laurberg\\
  }{%
    %list of supervisors
     Lone Leth Thomsen
  }{%
    \today % date of completion
  }%
}{%department and address
  \textbf{Department of Computer Science}\\
  Selma Lagerløfs Vej 300\\
  DK-9220 Aalborg Ø\\
  \href{http://www.cs.aau.dk}{http://www.cs.aau.dk}
}{% the abstract
Today the increase in \acs{CPU} speed comes in the form of additional cores. To utilise the potential of the \acs{CPU} programmers must make use of concurrency. We investigate trending concurrency models, and analyse both their runtime performance and characteristics. Java, Clojure and Scala with Akka, was chosen as representatives for \acl{TL}, \acl{STM}, and the Actor Model respectively. We contribute with a performance test of a $k$-means implementation and argue that the performance difference in most cases is negligible. The characteristics differs greatly, but none of the models are without problems. Each model shines in different scenarios, and so we contribute with recommendations for when to use each model. 
}
\end{titlingpage}