\makeatletter \@ifundefined{rootpath}{\input{../../setup/preamble.tex}}\makeatother
\worksheetstart{Project Introduction}{1}{September 5, 2014}{Kasper}{../../}
For many years Moore's law\cite{moore1965cramming} seemed to be true. The amount of transistors placed into an integrated circuit were approximately doubled every two years\andreas{Source!}, resulting in increasing processing speed. This increase has however stagnated\andreas{Source!} and processing power has mainly increased in the form of additional processing cores, as opposed to the speed of each of these processing cores. That is, computers have moved from having a \ac{CPU} consisting of one single core to one having multiple cores. 

The programming languages developed at the time while Moore's law still held, such as assembly and C, were closely related to the Von Neumann architecture. As a consequence, the programming model was designed for a synchronous execution. This introduces a challenge to how programs should be written to execute efficiently on a multi-core processor. A lot of effort has been invested into identifying suitable ways of programming against such architectures\andreas{Source?}. As of yet, no standard has been established. 

With the rise of the internet, the number of very large services supporting millions of users has increased. Furthermore many of these services present each user with dynamically generated content, limiting the usefulness of caching. Due to this, the need for scalable systems, and programming languages which suites the scalable approach arose. Previously, when scalability issues arose a company might invest in a new machine capable of handling the load, effectively discarding the investment of purchasing the old machine\cite[p. 2]{haller2007actors}. A more modern approach entails scaling the service over a set of low end machines, adding and removing machines as needed. However, the applications needs to be aware of this approach when developed. Frameworks for scaling in this fashion have been created\cite{haller2007actors}\cite{krohn2004building}\andreas{Denne sidste sætning står lidt alene nu}.

%Possibly something about the actor model

This project seeks to investigate different programming language constructs and technique, that seek to assist in scaling applications over multiple cores and preferable multiple machines.
%more stuff here

%Sprog
%Skrives som simpelt imperativt


 
\worksheetend
