\makeatletter \@ifundefined{rootpath}{\input{../../setup/preamble.tex}}\makeatother
\worksheetstart{Software Transactional Memory}{1}{November 7, 2013}{Andreas}{../../}
This chapter describes the concept of \acl{STM}. In \bsref{sec:stm_keyconcepts}, the key concepts of \ac{STM} is described. \bsref{sec:stm_issues} documents the concurrency issues introduces by \ac{STM}. \bsref{sec:stm_discussion} discusses the properties of \ac{STM}. Finally, in \bsref{sec:stm_eval} an evaluation of \ac{STM} as a concurrency model is conducted.

\section{Key Concepts}
\label{sec:stm_keyconcepts}
\kasper[inline]{Maybe stress the fact that the only different between \ac{TL} and \ac{STM} is in the way synchronization is done eg. locks vs transactions. Still shared memory leading to race conditions.}
\subsection{Transactions}
\ac{TM} is based on the concept of a transaction. A transaction is a finite sequence of machine instructions, executed by a single process, satisfying the properties: Atomicity, Consistency, and Isolation\cite[p. 2]{herlihy1993transactional}\cite[p. 14]{kalin2012clojure}. A transaction \ac{TM} can perform a group of memory-operations atomically\cite[48]{harris2005composable}. By utilizing the transactional approach, \ac{TM} serves as a concurrency control mechanism. By adhering to the aforementioned properties, \ac{TM} enables concurrent execution, while being semantically equivalent to serial execution of the same program.

To clarify how a transaction works, \bsref{} visualizes the flow of an example of a transaction.

\subsection{Software or Hardware}
In \cite{herlihy1993transactional} the Authors presents \ac{TM} as a straight forward extension to any multiprocessor cache-coherent protocol. Most \ac{HTM} proposals are based on straightforward modifications to standard multiprocessor cache-coherence protocols\cite{scott2011sync}. Due to the inflexible properties of existing hardware, i.e. fixed instruction set, \ac{TM} implemented on hardware is limited to existing instructions. This causes high implementation and verification costs\cite{cascaval2008software}. \ac{HTM} also exposes low level hardware details and limitations\cite{scott2011sync}, which preferably should be abstracted away from the programmer, to ensure semantic correctness of the transactions regardless of the hardware limitations. Due to this, a software based solution is presented by \cite{shavit1997software}. This solution, entirely implemented as software, is known as \acl{STM}. Being implemented as a software solution, \ac{STM} does not suffer the same issues as \ac{HTM}, but still provides the transactional semantics. Due to this, \ac{STM} is in interesting as a concurrency mechanism.

\subsection{\ac{STM}}
STM is a software system that provides programmers with a transactional model through a library or compiler interface\cite{scott2011sync}.

% Design criteria for STM (Thomas Stig fremlæggelse)
\subsection{Atomicity}

\subsection{Isolation level}
Strong vs Weak\\


%Transactions (Concept known from databases. HTM, STM)
%%ACI(D) properties

% STM
%%
%% Shared memory
%%% How to access shared data?

% Composition is possible

A number of programming languages and libraries have emerged to support TM. These include Clojure, .Net , Haskell. Java, C++\\
STM is easier to use: Pankratius, Rossbach\\
Speculative transactions\\

Exceptions\\
Interaction with code that cannot be transactionalized\\
Livelock\\
\section{Concurrency Issues}
\label{sec:stm_issues}
Structuring concurrent programs that seeks to provide better scalability and ease-of-use than conventional approaches based on locks and conditions.

\section{Discussion}
\label{sec:stm_discussion}
% Raceconditions. Transaction ensures atomicity (clojure got strong atomicity), and isolation. Due to this, the result of concurrent calculations will be the same, as if it was executed sequencially. 
% Deadlocks...

% Composition is possible in STM. Reference to Kaspers example about composition.

\section{Evaluation of Characteristics}
\label{sec:stm_eval}
% Optimistic (retry, not wait)

\worksheetend